---
title: "Intro to Stats"
author: "Aaron Weimann"
output: 
  html_document:
    theme: paper
    highlight: tango
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Simple plots in R with the ggplot2 library

Before start with the actual exercise, we first need to load the required libraries
```{r}
library(ggplot2)
library(data.table)
```

In addition, before, going into the idea of the data exploration proccess, one might question why it is importat at all to proyect numerical information into a 

## The Anscombe Quartet data as an example data set
```{r}
anscombe 
anscombe_m <- data.frame() # Create an empty data frame 

for(i in 1:4) { # convert dataset from wide to large format
  anscombe_m <- rbind(anscombe_m, data.frame(set = i, x = anscombe[,i], y = anscombe[,i+4]))
}
rm(i)
anscombe_m
```


## How to do simple plots in R 
```{r, message = F, warning= F}
ggplot(anscombe_m, aes(x, y)) + 
  geom_point(size=2, color="red", fill="orange", shape=21) + 
  geom_smooth(method="lm", fill=NA, fullrange=TRUE) + 
  facet_wrap(~paste0("Set ", set), ncol=2)
```

## Create an example dataset
```{r, message = F, warning= F}
Sigma <- matrix(c(10,4,3,2),2,2) # table of correlations for function mvrnorm
Sigma

d <- data.table(lugar = sample(c("España", "Mexico"), 400, replace = TRUE)) # variable1 = 2 countries
d[, grupo := rep(c("a", "b"), each = 200)]                                  # variable2 = 2 groups

# variable2 y 3: multivariate dist
set.seed(1) #  random seed to get the same results everytime we run this chunk of code

d[grupo == "a", 
  c("x", "y") :=  
    data.table(MASS::mvrnorm(n = 200, rep(1, 2), Sigma) *1.75)]             
d[grupo == "b", 
  c("x", "y") :=  
    data.table(MASS::mvrnorm(n = 200, rep(1, 2), Sigma) + 1.5)]
d[grupo == "b", y := y * 1.5 ]
d # Show the result
```

# Section 1: Simple plots 

## Histograms
```{r, message = F, warning= F}
ggplot(d, aes(x = y)) +
  geom_histogram(fill = "cornflowerblue", col = "gray80")   # use default number of bins in the histogram

ggplot(d, aes(x = y)) +
  geom_histogram(fill = "cornflowerblue", col = "gray80", 
                 bins = 64)                                 # use preset number of bins

```
## Quantile plots
```{r, message = F, warning= F}
ggplot(d, aes(sample = y)) +
  geom_qq(col = "cornflowerblue") +
  geom_qq_line(distribution = qnorm)
```

# Section 2: Plots with two variables 

## Bar plots
```{r, message = F, warning= F}
ggplot(d, aes(x = grupo, y = y)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") 
```
## Boxplots
```{r, message = F, warning= F}
ggplot(d, aes(x = grupo, y = y)) +
  geom_boxplot(fill = "cornflowerblue") 
```

# Section 4: Plots with continuous variables

```{r, message = F, warning= F}
ggplot(d, aes(x = x, y = y, shape = grupo)) +
  geom_point(col = "black", # color is fixed 
             size = 2) +
  theme_bw(base_size = 20)

ggplot(d, aes(x = x, y = y, col = grupo)) + # color according to the group
  geom_point(size = 2) +
  theme_bw(base_size = 20)
```

# Section 5: Representing simple models

```{r, message = F, warning= F}
ggplot(d, aes(x = x, y = y)) +
  geom_point() +
  stat_smooth() 

ggplot(d, aes(x = x, y = y)) +
  geom_point(aes(col = grupo)) +
  stat_smooth(method = "lm", col = "black")
```

# Section 6: Subset data and add an addtitional data series to an existing plot

```{r, message = F, warning= F}
ggplot(d, # Original data  
       aes(x = x, y = y, col = grupo)) +
  geom_point() +                                  # dot plots
  stat_smooth(method = "lm",                      # plotting an approximation of the point cloud
              formula = y ~ exp(x)) +             # specify approximating function
  
  # Second data set, 
  geom_line(data = d[, .(meanY = mean(y)),
              by = .(x = round(x), grupo, lugar)],
            
            aes(x = x, y = meanY),                # add a second data series
            linetype = 2, size = 1.5) +
 
  facet_grid(~ lugar,                             # have a facet for each group
             scales = "free_x") +
  scale_x_continuous(trans = "log2") +
  labs(x = "Una variable", y = "Una respuesta") +
  scale_color_brewer(palette = "Set1")
```
